Parsing Postfix log files

### 2008-07-10 doc/logparser-background.tex
Corrections from my last proofreading.

A snippet of \SQL{} provides a short example of the optimisation possible
using data from the database is determining which Postfix restrictions
reject the highest number of mails:

@@

    \item XXX REWRITE THIS PARAGRAPH\@.  The majority of programming
        languages have existing code available providing access to
        databases, so the majority of languages will be able to access the
        results obtained by the parser.  If this parser was written to be
        utilised as shared code there would be little or no difficulty in
        using it when writing in the programing language it is implemented
        in, but every other programing language wishing to utilise it would
        require an interface layer to be written.  This requirement would
        drastically reduce the viability of any project wishing to build
        upon the work of this project.

@@

Storing the results in a database will also increase the efficiency of
using those results, as the log files need only be parsed once; indeed the
results may be used by someone with no access to the original log files.

>>>>

A snippet of \SQL{} provides a short example of the optimisation possible
using data from the database: determining which Postfix restrictions reject
the highest number of mails:

@@

    \item It is already possible to access databases from the vast majority
        of programming languages, allowing a developer to access the
        data gathered using their preferred programming language, rather
        than being restricted to the language the parser is written in.  It
        is often possible to write an interface layer allowing code written
        in one language to be used in another language, but this greatly
        increases the effort required to use the parser.

@@

Storing the results in a database will also increase the efficiency of
using those results, as the log files need only be parsed once rather than
each time the data is used; indeed the results may be used by someone with
no access to the original log files.

### 2008-07-10 doc/logparser-database-schema.tex
Corrections from my last proofreading.

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in \sectionref{rules} but the fields are covered here
in \sectionref{rule attributes}.

The data saved from parsing the log files is also divided into two tables
as described below: connections and results.  The connections table
contains a row for every mail accepted and every connection where there was
a rejection; the individual fields will be described in
\sectionref{connections table}.  The results table has one or more rows per
row in the connections table; the fields will be covered in detail in
\sectionref{results table}.

An important but easily overlooked benefit of storing the rules in the
database is the connection between rules and results: if more information
is required when examining a result, the rules which produced the database
entries are available for inspection (and each result references the rule
which created it).  There is no ambiguity about which rule resulted in a
particular result, eliminating one potential source of confusion.

@@

Rules are discussed in detail in \sectionref{rules}, but the structure of
the rules table is covered here.  Rules are created by the user, not the
parser, and will not be modified by the parser (except for the hits and
hits\_total fields).  Rules parse the individual log lines, extracting data
to be saved in the connections and results tables, and specifying action to
take for that log line.

Each rule defines the following:

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line (e.g.\ a remote client connecting causes a connection
        line to be logged).  This field describes the action causing the
        log lines this rule matches.

@@

            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
                see~\cite{postfix-cleanup} for details of the processing
                performed by \daemon{cleanup}.

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Rule efficiency concerns are
        discussed in \sectionref{rule efficiency}, with the impact of
        compiling and caching \regexes{} covered in \sectionref{Caching
        each regex}.

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is:
        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
        \newline i.e.\ semi-colon or comma separated assignment statements,
        with the variable name on the left and the matching capture from
        the \regex{} on the right hand side.  The list of acceptable
        variable names is:

@@

        Additionally \texttt{child} and \texttt{pid} are used respectively
        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
        actions.

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data which is not present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This is intended for use with small amounts of
        data rather than large amounts in any one rule, so dealing with
        escape sequences was deemed unnecessary.} is permitted on the right
        hand side of the assignment.

@@

    \item [queueid] Specifies the matching capture from the \regex{} which
        gives the queueid, or zero if the log line does not contain a
        queueid.  Many log lines will not contain a queueid, e.g.\
        rejections logged before a mail has been accepted (a queueid will
        not have been allocated), or log lines which are not tied to one
        particular mail.

    \item [hits] This counter is maintained for every rule and incremented
        each time the rule successfully matches.  At the start of each run
        the program sorts the rules in descending order of hits, and at the
        end of the run updates every rule's hits.  Assuming that the
        distribution of log lines is reasonably consistent between log
        files, rules matching more commonly occurring log lines will be
        tried before rules matching less commonly occurring log lines,
        lowering the program's execution time.  Rule ordering for
        efficiency is discussed in \sectionref{rule ordering for
        efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: rules
        will be tried in order of priority, overriding hits.  This allows
        more specific rules to take precedence over more general rules
        (described in \sectionref{overlapping rules}).

@@

    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
        local server when receiving mail, the remote server when sending
        mail.

@@

    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
        server when receiving mail, the local server when sending mail.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \SMTP{} command, but it
        is quite easy to rectify this:

\label{logging helo}

        \begin{enumerate}

            \item Create \texttt{/etc/postfix/log\_helo.pcre}
                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}

            \item Modify \texttt{smtpd\_data\_restrictions} in
                \texttt{/etc/postfix/main.cf} to contain\newline
                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}

        \end{enumerate}

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there will not be a queueid
        associated with the mail for the first recipient, so that log line
        cannot be associated with the correct mail.  There is guaranteed to
        be a queueid when the DATA command has been reached, and thus it
        will be logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
        specifying a HELO-based restriction in
        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
        correctly.

        Logging the HELO hostname in this fashion also prevents the
        complication described in \sectionref{Mail deleted before delivery
        is attempted} from occurring when, but only in the case where there
        is a single recipient; in that case the recipient address will be
        logged also, but when there are multiple recipients none are
        logged.  It is also possible to warn for every recipient,
        preventing the complication in \sectionref{Mail deleted before
        delivery is attempted} entirely.

@@

Every log line where the associated rule has a \texttt{postfix\_action}
will have an entry in the results table, e.g.\ rejecting an \SMTP{}
command, delivering a mail, or bouncing a mail (see
\sectionref{postfix_action} for typical values of
\texttt{postfix\_action}).  Each row is associated with a single
connection, though there may be many results per connection.

@@

    \item [connection\_id] A reference to the row in the connections table
        this result is associated with.

    \item [rule\_id] A reference to the entry in the rules table which
        matched the log line and created this result.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \SMTP{} command --- a
        facility that is quite useful for testing new restrictions.  This
        field will be 1 if the log line parsed was a warning rather than a
        real rejection, or 0 for a real rejection.  This field should be
        ignored if the result is not a rejection.

    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
        general an \SMTP{} code is only present for a rejection or final
        delivery; results initially missing an \SMTP{} code will duplicate
        the \SMTP{} code of other results in the connection.  Some final
        delivery log lines do not contain an \SMTP{} code: in those cases
        the code is faked based on the success or failure represented by
        the log line.

    \item [sender] The sender's email address.  Because multiple mails may
        be delivered during one connection, there may be different sender
        addresses in the results for one connection; however there should
        not be different sender addresses in the results for one email.

>>>>

The database schema can be conceptually divided in two: the rules, which
are used to parse log files, and the data saved from the parsing of log
files.  Rules have the fields required to parse the log lines, extract data
to be saved, and the action to be executed; they also have several fields
which aid the user in understanding the meaning of the log lines parsed by
each rule.  The rules are described in detail in \sectionref{rules} but the
fields are covered in \sectionref{rule attributes}.

The data saved from parsing the log files is also divided into two tables
as described below: connections and results.  The connections table
contains a row for every mail accepted and every connection where there was
a rejection; the individual fields will be described in
\sectionref{connections table}.  There will be at least one row in the
results table for each row in the connections table; the fields will be
covered in detail in \sectionref{results table}.

An important but easily overlooked benefit of storing the rules in the
database is the connection between rules and results: if more information
is required when examining a result, the rule which produced the result is
available for inspection --- each result references the rule which created
it.  There is no ambiguity about which rule resulted in a particular
result, eliminating one potential source of confusion.

@@

Rules are discussed in detail in \sectionref{rules}, but the structure of
the rules table is covered here.  Rules are created by the user, not the
parser, and will not be modified by the parser (except for the hits and
hits\_total fields).  Rules parse the individual log lines, extracting data
to be saved in the connections and results tables, and specifying the
action to take for that log line.

The following attributes are defined for each rule:

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line (e.g.\ a remote client connecting causes a connection
        line to be logged).  This field describes the event causing the log
        lines this rule matches.

@@

            \item [PROCESSING] \daemon{cleanup} is processing a mail
                (see~\cite{postfix-cleanup} for details).

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Rule efficiency concerns are
        discussed in \sectionref{rule efficiency}, with the impact of
        caching compiled \regexes{} covered in \sectionref{Caching each
        regex}.

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is:
        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
        \newline i.e.\ semi-colon or comma separated assignment statements,
        with the variable name on the left and the matching capture from
        the \regex{} on the right hand side.  Data specified by
        \texttt{result\_cols} will be entered in the \texttt{results}
        table, while \texttt{connection\_cols} populates the
        \texttt{connection} table.  The list of acceptable variable names
        is:

@@

        Additionally \texttt{child} and \texttt{pid} are accepted and used
        respectively by the \texttt{TRACK}, \texttt{BOUNCE} and
        \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
        in the database.

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data which is not present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This feature is intended for use with small
        amounts of data rather than large amounts in any one rule, so
        dealing with escape sequences was deemed unnecessary.} is permitted
        on the right hand side of the assignment.

@@

    \item [queueid] Specifies the capturing field in the \regex{}
        containing the queueid, or zero if the log line does not contain a
        queueid.  Many log lines will not contain a queueid, e.g.\
        rejections logged before a mail has been accepted (a queueid will
        not have been allocated), or log lines which are not tied to one
        particular mail.

    \item [hits] This counter is maintained for every rule and incremented
        each time the rule successfully matches.  At the start of each run
        the program sorts the rules in descending order of hits, and at the
        end of the run updates every rule's hits field in the database.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, reducing the program's execution
        time.  Rule ordering for efficiency is discussed in
        \sectionref{rule ordering for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted, priority overrides hits.  This allows
        more specific rules to take precedence over more general rules
        (described in \sectionref{overlapping rules}).

@@

    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
        local address when receiving mail, the remote address when sending
        mail.

@@

    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
        address when receiving mail, the local address when sending mail.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \SMTP{} command, but it
        is quite easy to rectify this, as described in \sectionref{logging
        helo}.

@@

Every interesting\footnote{All log lines are interesting unless the action
of the rule which parses them is \texttt{INFO} or \texttt{IGNORE}, as
described in \sectionref{actions-in-detail}.} log line will have an entry
in the results table, e.g.\ rejecting an \SMTP{} command, delivering a
mail, or bouncing a mail (see \sectionref{postfix_action} for typical
values of \texttt{postfix\_action}).  Each row is associated with a single
connection, though there may be many results per connection.

@@

    \item [connection\_id] The id of the row in the connections table this
        result is associated with.

    \item [rule\_id] The id of the entry in the rules table which matched
        the log line and created this result.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \SMTP{} command --- a
        facility that is quite useful for testing new restrictions.  This
        field will be 1 if the log line parsed was a warning rather than a
        real rejection, or 0 for a real rejection.  This field should be
        ignored if the result is not a rejection, i.e.\ the action field of
        the associated rule is not \texttt{REJECTION}.

    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
        general an \SMTP{} code is only present for a rejection or final
        delivery; results initially missing an \SMTP{} code will duplicate
        the \SMTP{} code of other results in the connection.  Some final
        delivery log lines do not contain an \SMTP{} code: in those cases
        the code is specified by the rule, based on the success or failure
        represented by the log line.

    \item [sender] The sender's email address.  Because multiple mails may
        be delivered during one connection, there may be different sender
        addresses in the results for one connection; however there should
        not be different sender addresses in the results for one email.
        Mails sent over the same connection can be distinguished by their
        queueid.

### 2008-07-10 doc/logparser-limitations.tex
Corrections from my last proofreading.

@@

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
        2.5.}  Rectifying this has already been described in
        \sectionref{logging helo}.

>>>>

\label{logging helo}

@@

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
        2.5. XXX TEST THIS AGAIN.}  Rectifying this is relatively simple:
        create a restriction which is guaranteed to warn for every accepted
        mail, as follows:

            \item Create \texttt{/etc/postfix/log\_helo.pcre}
                containing:\newline \tab{}\texttt{/\^/~~~~WARN~Logging~HELO}

            \item Modify \texttt{smtpd\_data\_restrictions} in
                \texttt{/etc/postfix/main.cf} to contain\newline
                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there will not be a queueid
        associated with the mail for the first recipient, so the log line
        cannot be associated with the correct mail.  There is guaranteed
        to be a queueid when the DATA command has been reached, and thus it
        will be logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
        specifying a HELO-based restriction in
        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
        correctly.

        Logging the HELO hostname in this fashion also prevents the
        complication described in \sectionref{Mail deleted before delivery
        is attempted} from occurring, but only in the case where there is a
        single recipient; in that case the recipient address will be logged
        also, but when there are multiple recipients no addresses are
        logged.  It is also possible to warn for every recipient,
        preventing the complication in \sectionref{Mail deleted before
        delivery is attempted} entirely.

### 2008-07-10 doc/logparser-parsing-rules.tex
Corrections from my last proofreading.

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, followed by the problems caused by
overlapping rules, and techniques to detect and deal with such rules.  An
example rule and a log line it would match are provided, plus a description
of how the fields in the rule\footnote{The details of the table containing
the rules have already been described in \sectionref{rule attributes}.} are
used when matching a log line and subsequently performing the requested
action.  The section continues with a discussion of rule efficiency
concerns, referring to the graphs in \sectionref{graphs}, and finishes with
an explanation of the algorithm used to generate new \regexes{} from
unparsed lines.

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules have
        already have been tried and failed on the current log line.}  Any
        given rule will only be used to parse a subset of the log lines,
        and any given log line will only be parsed by a subset of the
        rules.

    \item The first matching rule wins: no further rules are tried against
        that log line, but there is a facility for ordering the rules so
        that more specific rules can be tried first.

@@

    \item Rule processing time is a linear function of the number of rules.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's \regex{} and determine if any overlap.  The
author has not attempted to do this: such a project by itself would
probably qualify for a PhD, and may involve solving the Halting
Problem~\cite{Wikipedia-halting-problem} and circumventing the
Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line may change between log
files, and the first matching rule wins.  Overlapping rules are frequently
a requirement, allowing a more specific rule to match some log lines and a
more general rule to match the majority, e.g.\ separating \SMTP{} delivery
to specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See \sectionref{rule efficiency} for more
        details of sorting the rules.}  Parse a number of log files using
        normal sorting, then dump a textual representation of the rules,
        connections and results tables.  Repeat with shuffled and reversed
        rules, starting with a fresh database.  If there are no overlapping
        rules the tables from each run will be identical; differences
        indicate overlapping rules.  Which rules overlap can be determined
        by examining the differences in the tables: each result contains a
        reference to the rule which created it, if the references differ
        between runs the two rules referenced in the differing records
        overlap.  Unfortunately this method cannot prove the absence of
        overlapping rules; it can detect overlapping rules, but only if
        there are log lines in the input files which match more than one
        rule.

>>>>

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, followed by the problems caused by
overlapping rules, and techniques to detect and deal with such overlaps.
An example rule and a log line it would match are provided, plus a
description of how the fields in the rule\footnote{The table containing the
rules has already been described in \sectionref{rule attributes}.} are used
when matching a log line and subsequently performing the requested action.
The section continues with a discussion of rule efficiency concerns, with
reference to the graphs in \sectionref{graphs}, and finishes with an
explanation of the algorithm used to generate new \regexes{} from unparsed
log lines.

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules have
        already have been tried and have failed on the current log line.}
        Any given rule will only be used to parse a subset of the log
        lines, and any given log line will only have a subset of the rules
        tried against it.

    \item The first matching rule wins: no further rules are tried against
        that log line, but there is a facility for prioritising the rules
        so that more specific rules can be tried first.

@@

    \item Rule processing time is a linear function of the number of rules.
        XXX IMPROVE THIS\@.

@@

XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?

The parser does not try to detect overlapping rules; that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line may change between log
files, and the first matching rule wins.  Overlapping rules are frequently
a requirement, allowing a more specific rule to match some log lines and a
more general rule to match the majority, e.g.\ separating \SMTP{} delivery
to specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Negative priorities may be useful
for catchall rules.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules (described in \sectionref{rule efficiency}).  Parse
        a number of log files using normal sorting, then dump a textual
        representation of the rules, connections and results tables.
        Repeat with shuffled and reversed rules, starting with a fresh
        database.  If there are no overlapping rules the tables from each
        run will be identical; differences indicate overlapping rules.
        Which rules overlap can be determined by examining the differences
        in the tables: each result contains a reference to the rule which
        created it, if the references differ between runs the two rules
        referenced in the differing results overlap.  Unfortunately this
        method cannot prove the absence of overlapping rules; it can detect
        overlapping rules, but only if there are log lines in the input
        files which match more than one rule.
